Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2023-02-09 analysis/paper/paper.Rmd
update bibliography

## Compile refs.bib using rbbt package ##
#cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
#cite_keys <- unlist(cite_keys, use.names = FALSE)
try(
  #rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
  rbbt::bbt_update_bib(
    here("analysis/paper/paper.Rmd"),
    here("analysis/paper/refs.bib"))
)
# ## end refs.bib ##

>>>>

## Compile refs.bib using rbbt package ##
#cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rmd_files <- list.files(path = here("analysis/paper/"), pattern = "*.Rmd")

try(
  #rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
  rbbt::bbt_update_bib(
    rmd_files,
    here("analysis/paper/refs.bib"))
)
# ## end refs.bib ##

### 2023-02-09 analysis/paper/paper.tex
update bibliography

\documentclass[utf8]{frontiers/frontiersSCNS}

@@

Statistical analysis was conducted in R version 4.2.0 (2022-04-22) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncient2013}{}}%
Adler, C. J., Dobney, K., Weyrich, L. S., Kaidonis, J., Walker, A. W., Haak, W., Bradshaw, C. J., Townsend, G., Soltysiak, A., Alt, K. W., et al. (2013). Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions. \emph{Nature Genetics} 45, 450--5, 455e1. doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.

\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentification1975}{}}%
Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science} 2, 187--197.

\leavevmode\vadjust pre{\hypertarget{ref-bernfeldAmylase1955}{}}%
Bernfeld, P. (1955). {``Amylases, {\(\alpha\)} and {\(\beta\)},''} in \emph{Methods in {Enzymology}} ({Academic Press}), 149--158. doi:\href{https://doi.org/10.1016/0076-6879(55)01021-5}{10.1016/0076-6879(55)01021-5}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cummingsMayanCalculus1997}{}}%
Cummings, L. S., and Magennis, A. (1997). {``A phytolith and starch record of food and grit in {Mayan} human tooth tartar,''} in \emph{The {State-of-the-Art} of {Phytoliths} in {Soils} and {Plants}}, eds. A. Pinilla, J. Juan-Tresserras, and M. J. Machado ({Spain}: {CSIC Press}).

\leavevmode\vadjust pre{\hypertarget{ref-doddsCarbohydrateRetention1988}{}}%
Dodds, M. W. J., and Edgar, W. M. (1988). The {Relationship Between Plaque pH}, {Plaque Acid Anion Profiles}, and {Oral Carbohydrate Retention After Ingestion} of {Several} '{Reference Foods}' by {Human Subjects}. \emph{Journal of Dental Research} 67, 861--865. doi:\href{https://doi.org/10.1177/00220345880670051301}{10.1177/00220345880670051301}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J. S., and Gang, D. R. (2018). Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}. \emph{Journal of Archaeological Science: Reports} 18, 509--515. doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-yatesOralMicrobiome2021}{}}%
Fellows Yates, J. A., Velsko, I. M., Aron, F., Posth, C., Hofman, C. A., Austin, R. M., Parker, C. E., Mann, A. E., Nägele, K., Arthur, K. W., et al. (2021). The evolution and changing ecology of the {African} hominid oral microbiome. \emph{Proceedings of the National Academy of Sciences} 118. doi:\href{https://doi.org/10.1073/pnas.2021655118}{10.1073/pnas.2021655118}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hardyRecoveringInformation2018}{}}%
Hardy, K., Buckley, S., and Copeland, L. (2018). Pleistocene dental calculus: {Recovering} information on {Paleolithic} food items, medicines, paleoenvironment and microbes. \emph{Evolutionary Anthropology: Issues, News, and Reviews} 27, 234--246. doi:\href{https://doi.org/10.1002/evan.21718}{10.1002/evan.21718}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hendyProteomicCalculus2018}{}}%
Hendy, J., Warinner, C., Bouwman, A., Collins, M. J., Fiddyment, S., Fischer, R., Hagan, R., Hofman, C. A., Holst, M., Chaves, E., et al. (2018). Proteomic evidence of dietary sources in ancient dental calculus. \emph{Proceedings. Biological Sciences} 285, 20180977. doi:\href{https://doi.org/10.1098/rspb.2018.0977}{10.1098/rspb.2018.0977}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-janeAnthologyStarch1994}{}}%
Jane, J.-L., Kasemsuwan, T., Leas, S., Zobel, H., and Robyt, J. F. (1994). Anthology of {Starch Granule Morphology} by {Scanning Electron Microscopy}. \emph{Starch - Stärke} 46, 121--129. doi:\href{https://doi.org/10.1002/star.19940460402}{10.1002/star.19940460402}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodRetention1991}{}}%
Kashket, S., Van Houte, J., Lopez, L. R., and Stocks, S. (1991). Lack of {Correlation Between Food Retention} on the {Human Dentition} and {Consumer Perception} of {Food Stickiness}. \emph{Journal of Dental Research} 70, 1314--1319. doi:\href{https://doi.org/10.1177/00220345910700100101}{10.1177/00220345910700100101}.

\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodParticles1996}{}}%
Kashket, S., Zhang, J., and Houte, J. V. (1996). Accumulation of {Fermentable Sugars} and {Metabolic Acids} in {Food Particles} that {Become Entrapped} on the {Dentition}. \emph{Journal of Dental Research}, 8.

@@

\leavevmode\vadjust pre{\hypertarget{ref-leonardDentalCalculus2015}{}}%
Leonard, C., Vashro, L., O'Connell, J. F., and Henry, A. G. (2015). Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists. \emph{Journal of Archaeological Science: Reports} 2, 449--457. doi:\href{https://doi.org/10.1016/j.jasrep.2015.03.009}{10.1016/j.jasrep.2015.03.009}.

\leavevmode\vadjust pre{\hypertarget{ref-lingstromStarchyFood1994}{}}%
Lingstrom, P., Birkhed, D., Ruben, J., and Arends, J. (1994). Effect of {Frequent Consumption} of {Starchy Food Items} on {Enamel} and {Dentin Demineralization} and on {Plaque pH} in situ. \emph{Journal of Dental Research} 73, 652--660. doi:\href{https://doi.org/10.1177/00220345940730031101}{10.1177/00220345940730031101}.

\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-middletonOpalPhytoliths1994}{}}%
Middleton, W. D., and Rovner, I. (1994). Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}. \emph{Journal of Archaeological Science} 21, 469--473. doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.

\leavevmode\vadjust pre{\hypertarget{ref-modiCalculusMethodologies2020}{}}%
Modi, A., Pisaneschi, L., Zaro, V., Vai, S., Vergata, C., Casalone, E., Caramelli, D., Moggi-Cecchi, J., Mariotti Lippi, M., and Lari, M. (2020). Combined methodologies for gaining much information from ancient dental calculus: Testing experimental strategies for simultaneously analysing {DNA} and food residues. \emph{Archaeological and Anthropological Sciences} 12, 10. doi:\href{https://doi.org/10.1007/s12520-019-00983-5}{10.1007/s12520-019-00983-5}.

\leavevmode\vadjust pre{\hypertarget{ref-Rhere}{}}%
Müller, K. (2020). \emph{Here: {A} simpler way to find your files}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nikitkovaStarchBiofilms2013}{}}%
Nikitkova, A. E., Haase, E. M., and Scannapieco, F. A. (2013). Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} {\(\alpha\)}-{Amylase Binding} to {Oral Streptococci}. \emph{Applied and Environmental Microbiology} 79, 416--423. doi:\href{https://doi.org/10.1128/AEM.02581-12}{10.1128/AEM.02581-12}.

\leavevmode\vadjust pre{\hypertarget{ref-pearceConcomitantDepositionStrontium1987}{}}%
Pearce, E. I. F., and Sissons, C. H. (1987). The {Concomitant Deposition} of {Strontium} and {Fluoride} in {Dental Plaque}. \emph{Journal of Dental Research} 66, 1518--1522. doi:\href{https://doi.org/10.1177/00220345870660100101}{10.1177/00220345870660100101}.

\leavevmode\vadjust pre{\hypertarget{ref-Rpatchwork}{}}%
Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-powerChimpCalculus2015}{}}%
Power, R. C., Salazar-Garcia, D. C., Wittig, R. M., Freiberg, M., and Henry, A. G. (2015). Dental calculus evidence of {Tai Forest Chimpanzee} plant consumption and life history transitions. \emph{Scientific Reports} 5, 15161. doi:\href{https://doi.org/10.1038/srep15161}{10.1038/srep15161}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-powerRepresentativenessDental2021}{}}%
Power, R. C., Wittig, R. M., Stone, J. R., Kupczik, K., and Schulz-Kornas, E. (2021). The representativeness of the dental calculus dietary record: Insights from {Taï} chimpanzee faecal phytoliths. \emph{Archaeological and Anthropological Sciences} 13, 104. doi:\href{https://doi.org/10.1007/s12520-021-01342-z}{10.1007/s12520-021-01342-z}.

\leavevmode\vadjust pre{\hypertarget{ref-proctorSpatialGradient2018}{}}%
Proctor, D. M., Fukuyama, J. A., Loomer, P. M., Armitage, G. C., Lee, S. A., Davis, N. M., Ryder, M. I., Holmes, S. P., and Relman, D. A. (2018). A spatial gradient of bacterial diversity in the human oral cavity shaped by salivary flow. \emph{Nature Communications} 9, 681. doi:\href{https://doi.org/10.1038/s41467-018-02900-1}{10.1038/s41467-018-02900-1}.

\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.

\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodPathways2017}{}}%
Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K. (2017). Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus. \emph{American Journal of Physical Anthropology} 162, 71--83. doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rbroom}{}}%
Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}.

\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}}%
Scannapieco, F. A., Torres, G., and Levine, M. J. (1993). Salivary {\(\alpha\)}-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine} 4, 301--307.

@@

\leavevmode\vadjust pre{\hypertarget{ref-simon-soroOralGeography2013}{}}%
Simón-Soro, A., Tomás, I., Cabrera-Rubio, R., Catalan, M. D., Nyvad, B., and Mira, A. (2013). Microbial geography of the oral cavity. \emph{Journal of Dental Research} 92, 616--621. doi:\href{https://doi.org/10.1177/0022034513488119}{10.1177/0022034513488119}.

\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationPlaque1991}{}}%
Sissons, C. H., Cutress, T. W., Hoffman, M. P., and Wakefield, J. S. J. (1991). A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}. doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.

\leavevmode\vadjust pre{\hypertarget{ref-tanCalculusUltrastructure2004}{}}%
Tan, B. T. K., Gillam, D. G., Mordan, N. J., and Galgut, P. N. (2004a). A preliminary investigation into the ultrastructure of dental calculus and associated bacteria. \emph{Journal of Clinical Periodontology} 31, 364--369. doi:\href{https://doi.org/10.1111/j.1600-051X.2004.00484.x}{10.1111/j.1600-051X.2004.00484.x}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M., et al. (2014a). Direct evidence of milk consumption from ancient human dental calculus. \emph{Scientific Reports} 4, 7104. doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.

\leavevmode\vadjust pre{\hypertarget{ref-warinnerPathogensHost2014}{}}%
Warinner, C., Rodrigues, J. F., Vyas, R., Trachsel, C., Shved, N., Grossmann, J., Radini, A., Hancock, Y., Tito, R. Y., Fiddyment, S., et al. (2014b). Pathogens and host immunity in the ancient human oral cavity. \emph{Nature Genetics} 46, 336--44. doi:\href{https://doi.org/10.1038/ng.2906}{10.1038/ng.2906}.

\leavevmode\vadjust pre{\hypertarget{ref-wesolowskiEvaluatingMicrofossilContent2010}{}}%
Wesolowski, V., Ferraz Mendonça de Souza, S. M., Reinhard, K. J., and Ceccantini, G. (2010). Evaluating microfossil content of dental calculus from {Brazilian} sambaquis. \emph{Journal of Archaeological Science} 37, 1326--1338. doi:\href{https://doi.org/10.1016/j.jas.2009.12.037}{10.1016/j.jas.2009.12.037}.

>>>>

\documentclass[utf8]{../templates/frontiersSCNS}

@@

Statistical analysis was conducted in R version 4.2.2 Patched (2022-11-10 r83330) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncient2013}{}}%
Adler, C. J., Dobney, K., Weyrich, L. S., Kaidonis, J., Walker, A. W., Haak, W., Bradshaw, C. J., Townsend, G., Sołtysiak, A., Alt, K. W., et al. (2013). Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions. \emph{Nat Genet} 45, 450--5, 455e1. doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.

\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentification1975}{}}%
Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{J Archaeol Sci} 2, 187--197.

\leavevmode\vadjust pre{\hypertarget{ref-bernfeldAmylase1955}{}}%
Bernfeld, P. (1955). {``Amylases, α and β,''} in \emph{Methods in {Enzymology}} ({Academic Press}), 149--158. doi:\href{https://doi.org/10.1016/0076-6879(55)01021-5}{10.1016/0076-6879(55)01021-5}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cummingsMayanCalculus1997}{}}%
Cummings, L. S., and Magennis, A. (1997). {``A phytolith and starch record of food and grit in {Mayan} human tooth tartar,''} in \emph{The {State-of-the-Art} of {Phytoliths} in {Soils} and {Plants}}, eds. A. Pinilla, J. Juan-Tresserras, and M. J. Machado ({Spain}: {CSIC Press}). Available at: \url{https://books.google.com?id=j66CDVfVhwEC}.

\leavevmode\vadjust pre{\hypertarget{ref-doddsCarbohydrateRetention1988}{}}%
Dodds, M. W. J., and Edgar, W. M. (1988). The {Relationship Between Plaque pH}, {Plaque Acid Anion Profiles}, and {Oral Carbohydrate Retention After Ingestion} of {Several} '{Reference Foods}' by {Human Subjects}. \emph{J Dent Res} 67, 861--865. doi:\href{https://doi.org/10.1177/00220345880670051301}{10.1177/00220345880670051301}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J. S., and Gang, D. R. (2018). Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}. \emph{J Archaeol Sci Rep} 18, 509--515. doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-yatesOralMicrobiome2021}{}}%
Fellows Yates, J. A., Velsko, I. M., Aron, F., Posth, C., Hofman, C. A., Austin, R. M., Parker, C. E., Mann, A. E., Nägele, K., Arthur, K. W., et al. (2021). The evolution and changing ecology of the {African} hominid oral microbiome. \emph{PNAS} 118. doi:\href{https://doi.org/10.1073/pnas.2021655118}{10.1073/pnas.2021655118}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hardyRecoveringInformation2018}{}}%
Hardy, K., Buckley, S., and Copeland, L. (2018). Pleistocene dental calculus: {Recovering} information on {Paleolithic} food items, medicines, paleoenvironment and microbes. \emph{Evol Anthropol} 27, 234--246. doi:\href{https://doi.org/10.1002/evan.21718}{10.1002/evan.21718}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hendyProteomicCalculus2018}{}}%
Hendy, J., Warinner, C., Bouwman, A., Collins, M. J., Fiddyment, S., Fischer, R., Hagan, R., Hofman, C. A., Holst, M., Chaves, E., et al. (2018). Proteomic evidence of dietary sources in ancient dental calculus. \emph{Proc Biol Sci} 285, 20180977. doi:\href{https://doi.org/10.1098/rspb.2018.0977}{10.1098/rspb.2018.0977}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-janeAnthologyStarch1994}{}}%
Jane, J.-L., Kasemsuwan, T., Leas, S., Zobel, H., and Robyt, J. F. (1994). Anthology of {Starch Granule Morphology} by {Scanning Electron Microscopy}. \emph{Starch/Stärke} 46, 121--129. doi:\href{https://doi.org/10.1002/star.19940460402}{10.1002/star.19940460402}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodRetention1991}{}}%
Kashket, S., Van Houte, J., Lopez, L. R., and Stocks, S. (1991). Lack of {Correlation Between Food Retention} on the {Human Dentition} and {Consumer Perception} of {Food Stickiness}. \emph{J Dent Res} 70, 1314--1319. doi:\href{https://doi.org/10.1177/00220345910700100101}{10.1177/00220345910700100101}.

\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodParticles1996}{}}%
Kashket, S., Zhang, J., and Houte, J. V. (1996). Accumulation of {Fermentable Sugars} and {Metabolic Acids} in {Food Particles} that {Become Entrapped} on the {Dentition}. \emph{J Dent Res}, 8.

@@

\leavevmode\vadjust pre{\hypertarget{ref-leonardDentalCalculus2015}{}}%
Leonard, C., Vashro, L., O'Connell, J. F., and Henry, A. G. (2015). Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists. \emph{J Archaeol Sci Rep} 2, 449--457. doi:\href{https://doi.org/10.1016/j.jasrep.2015.03.009}{10.1016/j.jasrep.2015.03.009}.

\leavevmode\vadjust pre{\hypertarget{ref-lingstromStarchyFood1994}{}}%
Lingstrom, P., Birkhed, D., Ruben, J., and Arends, J. (1994). Effect of {Frequent Consumption} of {Starchy Food Items} on {Enamel} and {Dentin Demineralization} and on {Plaque pH} in situ. \emph{J Dent Res} 73, 652--660. doi:\href{https://doi.org/10.1177/00220345940730031101}{10.1177/00220345940730031101}.

\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}. Available at: \url{https://github.com/benmarwick/rrtools}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-middletonOpalPhytoliths1994}{}}%
Middleton, W. D., and Rovner, I. (1994). Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}. \emph{J Archaeol Sci} 21, 469--473. doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.

\leavevmode\vadjust pre{\hypertarget{ref-modiCalculusMethodologies2020}{}}%
Modi, A., Pisaneschi, L., Zaro, V., Vai, S., Vergata, C., Casalone, E., Caramelli, D., Moggi-Cecchi, J., Mariotti Lippi, M., and Lari, M. (2020). Combined methodologies for gaining much information from ancient dental calculus: Testing experimental strategies for simultaneously analysing {DNA} and food residues. \emph{Archaeol Anthropol Sci} 12, 10. doi:\href{https://doi.org/10.1007/s12520-019-00983-5}{10.1007/s12520-019-00983-5}.

\leavevmode\vadjust pre{\hypertarget{ref-Rhere}{}}%
Müller, K. (2020). \emph{Here: {A} simpler way to find your files}. Available at: \url{https://CRAN.R-project.org/package=here}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nikitkovaStarchBiofilms2013}{}}%
Nikitkova, A. E., Haase, E. M., and Scannapieco, F. A. (2013). Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}. \emph{Appl. Environ. Microbiol.} 79, 416--423. doi:\href{https://doi.org/10.1128/AEM.02581-12}{10.1128/AEM.02581-12}.

\leavevmode\vadjust pre{\hypertarget{ref-pearceConcomitantDepositionStrontium1987}{}}%
Pearce, E. I. F., and Sissons, C. H. (1987). The {Concomitant Deposition} of {Strontium} and {Fluoride} in {Dental Plaque}. \emph{J Dent Res} 66, 1518--1522. doi:\href{https://doi.org/10.1177/00220345870660100101}{10.1177/00220345870660100101}.

\leavevmode\vadjust pre{\hypertarget{ref-Rpatchwork}{}}%
Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}. Available at: \url{https://CRAN.R-project.org/package=patchwork}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-powerChimpCalculus2015}{}}%
Power, R. C., Salazar-Garcia, D. C., Wittig, R. M., Freiberg, M., and Henry, A. G. (2015). Dental calculus evidence of {Tai Forest Chimpanzee} plant consumption and life history transitions. \emph{Sci Rep} 5, 15161. doi:\href{https://doi.org/10.1038/srep15161}{10.1038/srep15161}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-powerRepresentativenessDental2021}{}}%
Power, R. C., Wittig, R. M., Stone, J. R., Kupczik, K., and Schulz-Kornas, E. (2021). The representativeness of the dental calculus dietary record: Insights from {Taï} chimpanzee faecal phytoliths. \emph{Archaeol Anthropol Sci} 13, 104. doi:\href{https://doi.org/10.1007/s12520-021-01342-z}{10.1007/s12520-021-01342-z}.

\leavevmode\vadjust pre{\hypertarget{ref-proctorSpatialGradient2018}{}}%
Proctor, D. M., Fukuyama, J. A., Loomer, P. M., Armitage, G. C., Lee, S. A., Davis, N. M., Ryder, M. I., Holmes, S. P., and Relman, D. A. (2018). A spatial gradient of bacterial diversity in the human oral cavity shaped by salivary flow. \emph{Nat Commun} 9, 681. doi:\href{https://doi.org/10.1038/s41467-018-02900-1}{10.1038/s41467-018-02900-1}.

\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing} Available at: \url{https://www.R-project.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodPathways2017}{}}%
Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K. (2017). Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus. \emph{Am J Phys Anthropol} 162, 71--83. doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rbroom}{}}%
Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}. Available at: \url{https://CRAN.R-project.org/package=broom}.

\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}}%
Scannapieco, F. A., Torres, G., and Levine, M. J. (1993). Salivary α-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine} 4, 301--307.

@@

\leavevmode\vadjust pre{\hypertarget{ref-simonsoroOralGeography2013}{}}%
Simón-Soro, A., Tomás, I., Cabrera-Rubio, R., Catalan, M. D., Nyvad, B., and Mira, A. (2013). Microbial geography of the oral cavity. \emph{J Dent Res} 92, 616--621. doi:\href{https://doi.org/10.1177/0022034513488119}{10.1177/0022034513488119}.

\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationPlaque1991}{}}%
Sissons, C. H., Cutress, T. W., Hoffman, M. P., and Wakefield, J. S. J. (1991). A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{J Dent Res}. doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.

\leavevmode\vadjust pre{\hypertarget{ref-tanCalculusUltrastructure2004}{}}%
Tan, B. T. K., Gillam, D. G., Mordan, N. J., and Galgut, P. N. (2004a). A preliminary investigation into the ultrastructure of dental calculus and associated bacteria. \emph{J Clin Periodontol} 31, 364--369. doi:\href{https://doi.org/10.1111/j.1600-051X.2004.00484.x}{10.1111/j.1600-051X.2004.00484.x}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M., et al. (2014a). Direct evidence of milk consumption from ancient human dental calculus. \emph{Sci Rep} 4, 7104. doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.

\leavevmode\vadjust pre{\hypertarget{ref-warinnerPathogensHost2014}{}}%
Warinner, C., Rodrigues, J. F., Vyas, R., Trachsel, C., Shved, N., Grossmann, J., Radini, A., Hancock, Y., Tito, R. Y., Fiddyment, S., et al. (2014b). Pathogens and host immunity in the ancient human oral cavity. \emph{Nat Genet} 46, 336--44. doi:\href{https://doi.org/10.1038/ng.2906}{10.1038/ng.2906}.

\leavevmode\vadjust pre{\hypertarget{ref-wesolowskiEvaluatingMicrofossilContent2010}{}}%
Wesolowski, V., Ferraz Mendonça de Souza, S. M., Reinhard, K. J., and Ceccantini, G. (2010). Evaluating microfossil content of dental calculus from {Brazilian} sambaquis. \emph{J Archaeol Sci} 37, 1326--1338. doi:\href{https://doi.org/10.1016/j.jas.2009.12.037}{10.1016/j.jas.2009.12.037}.
